Fast response of mesophyll conductance to light availability allows shade leaves to take advantage of sunflecks.

### 2015-08-19 manuscript/manuscript_wtc3.Rmd
made change
manuscript edits
made changes to sup figure alpha of points

added inset to acc plot

add more cites

```{r, echo=FALSE}
airvarsfig <- "1"
parfig <- "2"
acifig <- "3"
waterusefig <- "4"
gebarfig <- "5"
agmgsfig <- "6"
deltafig <- "7"
ciccfig <- "8"

get1fig <- "S1"
get2fig <- "S2"
accfig <- "S3"
```

@@

where the term Hp~w~g represents the gravitational potential assumed to be minimal in this experiment due to small tree size [@whitehead1998regulation]. Leaf level instantaneous water use efficiency (WUE~i~) was calculated as leaf photosynthesis divided by transpiration. The g~1~ parameter was estimated from ITE to VPD response curves by fitting a rearranged optimal stomatal conductance model for ITE [@medlyn2012reconciling] using non-linear regression, where K=0.5 [see @duursma2013near].

@@

## Leaf resource distribution
Across all measurement campaigns 1 m canopy integrated PPFD was significantly different between sun and shade leaves (P<0.001) and PPFD was reduced by 78% in the shade (Figure `r parfig`). Leaf-specific hydraulic conductance was not different across sun and shade leaves (Table 1).  This was the result of neither $\Psi_{pd}$, $\Psi_l$ (Table 1) nor E (Table 2) differing between leaf types. Leaf N, on an area basis, was approximately 20% higher in sun leaves than in shade leaves (Table1). Leaf mass per area was not different between leaf types. No effect of the warming treatment was detected with  PPFD, $\Psi_{pd}$, $\Psi_l$, K, E, N~area~ or LMA either within or across leaf types.

## Photosynthetic capacity and leaf photosynthesis rates 
The photosynthetic parameters J~max~ and Vc~max~ were both significantly higher in sun than shade leaves (Table 1), as evident from AC~i~ curves from each chamber (Figure `r acifig`a). Within leaf types, no effect of the warming treatment was detected on either parameter. Within individual chambers Vc~max~ was positively related to leaf N~area~ across leaf types and temperature treatments (P = 0.01, Figure `r acifig`b). Additionally, leaf N~area~ was positively related to A across gas exchange campaigns and leaf types (P < 0.001, Figure `r acifig`c). As g~m~ (results below) is included in the estimates of J~max~ and Vc~max~ with conventional AC~i~, AC~c~ curves were also simulated to determine if treatment differences in AC~i~ parameters where instead the result of differences in g~m~. When accounting for the effect of g~m~, chloroplastic Vc~max~ and J~max~ were still higher in sun leaves (Figure `r accfig`).

\
Mean leaf photosynthesis rates were significantly higher in sun compared to shade leaves (+23%), when measured at their local light environment (Table 2). Following an increase in available light, A of shade leaves was significantly greater than both shade leaves at low light and sun leaves (P < 0.001). No effect of the warming treatment was detected on rates of A of sun leaves or shade leaves at low or high light. Photosynthesis within leaf types and treatments was similar through time and across the range of leaf temperatures measured (Figure `r get1fig`a).

## Leaf water use efficiency
Leaf WUE~i~ was significantly greater in sun leaves than in shade leaves at low light (P = 0.001). Following an increase in available light, WUE~i~ of shade leaves did not differ from shade leaves at low light and was still significantly lower than sun leaves (P < 0.001). WUE~i~ in sun leaves was reduced in the warming treatment but no effect was noted in shade leaves with low or high light (Table 2). For all leaf types and light treatments there was a strong response of WUE~i~ to VPD, and individual data points broadly corresponded to response curves from the optimal WUE~i~ model (Figure `r waterusefig`a). The mean estimated g~1~ for sun leaves was 1.51&pm;0.11 and for shade leaves with low and high light was 2.59&pm;0.12 and 2.74&pm;0.04. Within leaf types and light treatments the response of VPD to leaf temperature was similar across all measurement campaigns (Figure `r get2fig`a)

@@

## Leaf physiology
Mean g~s~ was significantly higher in shade compared to sun leaves (+18%) under their local light environment (P = 0.004). Increases in g~s~ were postively correlated to increases in A in sun leaves, however, g~s~ and A were not correlated in shade leaves under low light (Figure `r agmgsfig`a). Following an increase in available light, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves ( Figure `r gebarfig`a). With increased light availability, both increases in A and g~s~ in shade leaves exhibited similar behavior as observed in sun leaves.

\
The observed carbon isotope discrimination was positively correlated with C~i~:C~a~ for all leaf types (P < 0.001) with larger $\Delta$~o~ detected for sun leaves and shade leaves at high light (Figure `r deltafig`). Here, the carbon isotope discrimination associated with g~m~ accounted for the majority of $\Delta$~o~ (69.7&pm;0.4 %) and varied little across measurement temperatures, leaf types, or warming treatments.

@@

\
The combination of a non-limiting supply of leaf K and more open stomata lead to a significant increase in C~i~ in shade compared to sun leaves (Figure `r ciccfig`a). As increases in A in shade leaves were associated with higher rates of g~s~ the C~i~ was similar than under low light but still greater than sun leaves. As the drop in CO~2~ for intercellular spaces to the cholorplast, C~i~-C~c~, measures the coordination between g~m~ and A [@caemmerer2014temperature] an increase in this drawdown was detected in shade leaves at high light (Figure `r ciccfig`b,c).  This was a direct result of the increase in g~m~ and its subsequent direct effect on A, when compared to both sun and shade leaves that are acclimated to their local light environment. CO~2~ drawdown from C~a~ to C~i~ and C~i~ to C~c~ were both relatively stable across the range of temperatures measured and gas exchange campaigns (Figure `r get2fig`c and `r get1fig`c, respectively).

@@

\
Photosynthesis in trees is also limited by the ability to supply water to the upper canopy. Using a two-leaf model Peltoniemi et al. [-@peltoniemi2012co] theorizes that optimal N distribution will be proportional to light distribution only if leaf K is also optimally distributed. In this study, variation in leaf N distribution and A rates were not associated with subsequent changes in leaf K between sun and shade leaves. This is especially meaningful in trees as foliage generally operates at greater water limitations in the upper canopy. As both sun and shade leaves were acclimated to their local light environment this sub-optimal distribution of water supply has the potential to diminish the C return for the whole canopy.

\
The inefficient distribution of resources led to decreased instantaneous WUE in shade leaves throughout the experiment.  Additionally, consistently higher leaf $\delta$^13^C in shade leaves suggests that this pattern was constant across leaf lifespans.  From a canopy perspective this pattern in WUE appears to be detrimental to C gain as A in sun leaves was characterized by low rates of g~s~ and low C~i~ compared to similar Eucalyptus species [see @drake2014capacity].  As whole canopy C gain integrates the efficiency of all leaves, this begs the question of why shade leaves maintained a lower WUE while the photosynthetic efficiency of sun leaves appears at least partially constrained by water supply.

@@

\
This behaviour may be explained by the response of shade leaves to increases in light availability. Contrary to hypothesis, shade leaves were able to increase A to rates equivalent or even outperforming sun leaves under identical light intensity. It is possible that sustaining higher g~s~, at the cost of larger transpiration losses, is a strategy to increase the utilization efficiency of sunflecks. These results should not be entirely unexpected as shade leaves experience interchanging periods of sun and shade, despite lower mean PPFD [@pearcy1990sunflecks]. Unfortunately, empirical studies that assess photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@leakey2002relative] and understory plants [@chazdon1991importance; @brantley2009contribution]. Our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy.

@@

\
Furthermore, our findings have important implications for process based models that predict canopy C gain from rates of leaf photosynthesis. Although the distributions of resources, such as N and water, define leaf photosynthetic capacity it is the physiological behaviour of individual leaves which actually determine C gain. The observed physiological behaviour leaves in the shade, at both low and high irradiance, suggests that current theories of leaf optimal behaviour are incomplete.  Furthermore, g~m~ must now be considered as a dynamic process in tree growth models that cannot be simply parameterized.  Additional empirical data are needed to determine both the mechanisms and the capacity of g~m~ to rapidly increase CO~2~ drawdown. To improve our ability to predict whole canopy C gain future research should prioritize the incorporation of both sun and shade leaf physiology, which may be optimized differently.

@@

**Table 1**. *Eucalyptus tereticornis* leaf traits and the distribution of resources between full sun and shade leaves under ambient and elevated temperature treatments. Leaf mass per area, leaf Nitrogen~area~, $\delta$^13^C, $\Psi$~pd~, $\Psi$~l~ and K values represent treatment mean (&pm; 1 standard error) across measurement campaigns (n=6). Values of Vc~max~ and J~max~ are treatment mean(standard error) from AC~i~ curves measured in each chamber at saturating light. Different letters represent significant differences between treatments. The P value represents the overall difference between leaf types and warming treatments.

@@

\
**Figure 4**. (a) Response of instantaneous water use efficiency (ITE) to leaf vapor pressure deficit (VPD) for sun leaves and shade leaves at both low and high light with elevated and ambient temperature treatments. (b) The relationship between leaf $\delta$^13^C and leaf nitrogen~area~ for sun leaves and shade leaves at low light. For (a) VPD is the leaf to air pressure difference inside the gas exchange cuvette and lines represent predictions from the optimal ITE model with a g1 value for each leaf type and treatment. For (b) the dashed line represents the significant linear model fit for all leaves with a marginal and conditional r^2^ of 0.41 and 0.45, respectively.

@@

\
**Figure 6**. The response of leaf photosynthesis rates to stomatal conductance (a) and mesophyll conductance (b) for sun leaves and shade leaves at both low and high light with elevated and ambient temperature treatments.  The response of shade leaf physiology to high light was recorded once both photosynthesis and stomatal conductance were stable in the gas exchange leaf cuvette (ca. 25min). Lines represent either smoothed regressions from a generalized additive model fit (a) or linear model fits (b). Grey areas are approximately 95% confidence intervals from the mean.

@@

\
**Figure S1**. Response of photosynthesis (a), mesophyll conductance (b) and C~i~-C~c~ to leaf temperature for sun leaves and shade leaves at low and high light.  Shaded symbols represents each monthly measurement campaign.

\
**Figure S2**. Response of vapor pressure deficit (a), stomatal conductance (b) and C~a~-C~i~ to leaf temperature for sun leaves and shade leaves at low and high light.  Shaded symbols represents each monthly measurement campaign.

\
**Figure S3**. Photosynthetic CO~2~ response (AC~c~) curves for sun and shade leaves at elevated and ambient temperature treatments. C~c~ values were predicted with mesophyll conductance, thus curves represent chloroplastic photosynthetic parameters at 25°C and at saturating light (1800 $\mu$mols m^-1^ s^-1^).

@@

@@

```{r Agmgs, fig.width=6, fig.height=8, echo=FALSE,warning=FALSE,dev="png", dpi=400, dev.args=list(type="cairo")}
source("master_scripts/Agmgs.R")
```

@@

# Supporting Information {.unnumbered}
```{r gasex_temp1, fig.width=6, fig.height=10, echo=FALSE,warning=FALSE,dev="png", dpi=400, dev.args=list(type="cairo")}
source("master_scripts/gasex_temp.R")
```  
**Figure S1**.

```{r gasex_temp2, fig.width=6, fig.height=10, echo=FALSE,warning=FALSE,dev="png", dpi=400, dev.args=list(type="cairo")}
source("master_scripts/gasex_temp2.R")
```  
**Figure S2**.

@@

>>>>

```{r, echo=FALSE}
airvarsfig <- "1"
parfig <- "2"
acifig <- "3"
agmgsfig <- "4"
gebarfig <- "5"
waterusefig <- "6"
deltafig <- "7"
ciccfig <- "8"

accfig <- "S1"
get1fig <- "S2"
get2fig <- "S3"

```

@@

where the term Hp~w~g represents the gravitational potential assumed to be minimal in this experiment due to small tree size [@whitehead1998regulation]. Leaf level instantaneous water use efficiency (~i~WUE) was calculated as leaf photosynthesis divided by transpiration. The g~1~ parameter was estimated from ITE to VPD response curves by fitting a rearranged optimal stomatal conductance model for ITE [@medlyn2012reconciling] using non-linear regression, where K=0.5 [see @duursma2013near].

@@

## Leaf resource distribution
Across all measurement campaigns PPFD was significantly different between sun and shade leaves (P<0.001) and PPFD was reduced by 78% in the shade (Figure `r parfig`). Leaf-specific hydraulic conductance was similar across sun and shade leaves (Table 1). This was because $\Psi_{pd}$, $\Psi_l$ (Table 1) and E (Table 2) did not differ between leaf types. Leaf N on an area basis (N~a~) was approximately 20% higher in sun leaves than in shade leaves (Table1). Leaf mass per area was not different between leaf types. No effect of the warming treatment was detected with  PPFD, $\Psi_{pd}$, $\Psi_l$, K, E, N~area~ or LMA either within or across leaf types.

## Photosynthetic capacity and leaf photosynthesis rates 
The photosynthetic parameters J~max~ and Vc~max~ were both significantly higher in sun than shade leaves (Table 1), as evident from AC~i~ curves from each chamber (Figure `r acifig`a). Within leaf types, no effect of the warming treatment was detected on either parameter. Within individual chambers Vc~max~ was positively related to leaf N~a~ across leaf types and temperature treatments (P = 0.01, Figure `r acifig`b). As g~m~ (results below) is included in the estimates of J~max~ and Vc~max~ with conventional AC~i~, AC~c~ curves were also simulated to determine if treatment differences in AC~i~ parameters where instead the result of differences in g~m~. When accounting for the effect of g~m~, chloroplastic Vc~max~ and J~max~ were still higher in sun leaves (Figure `r accfig`).

\
Mean leaf photosynthesis rates were significantly higher in sun compared to shade leaves (+23%), when measured at their local light environment (Table 2). Additionally, leaf N~area~ was positively related to A across gas exchange campaigns and leaf types (P < 0.001, Figure `r acifig`c). Following an increase in available light, A of shade leaves was significantly greater than both shade leaves at low light and sun leaves (P < 0.001). No effect of the warming treatment was detected on rates of A of sun leaves or shade leaves at low or high light. Photosynthesis within leaf types and treatments was similar through time and across the range of leaf temperatures measured (Figure `r get1fig`a).

## Stomatal conductance and leaf water use efficiency
Mean g~s~ was significantly higher in shade compared to sun leaves (+18%) under their local light environment (P = 0.004). Increases in g~s~ were postively correlated to increases in A in sun leaves, however, g~s~ and A were not correlated in shade leaves under low light (Figure `r agmgsfig`a). Following an increase in available light, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves ( Figure `r gebarfig`a). With increased light availability, both increases in A and g~s~ in shade leaves exhibited similar behavior as observed in sun leaves.

\
Leaf ~i~WUE was significantly greater in sun leaves than in shade leaves at low light (P = 0.001). Following an increase in available light, ~i~WUE of shade leaves did not differ from shade leaves at low light and was still significantly lower than sun leaves (P < 0.001). ~i~WUE in sun leaves was reduced in the warming treatment but no effect was noted in shade leaves with low or high light (Table 2). For all leaf types and light treatments there was a strong response of ~i~WUE to VPD, and individual data points broadly corresponded to response curves from the optimal ~i~WUE model (Figure `r waterusefig`a). The mean estimated g~1~ for sun leaves was 1.51&pm;0.11 and for shade leaves with low and high light was 2.59&pm;0.12 and 2.74&pm;0.04. Within leaf types and light treatments the response of VPD to leaf temperature was similar across all measurement campaigns (Figure `r get2fig`a)

@@

## Leaf carbon isotope discrimination and mesophyll conductance
The observed carbon isotope discrimination was positively correlated with C~i~:C~a~ for all leaf types (P < 0.001) with larger $\Delta$~o~ detected for sun leaves and shade leaves at high light (Figure `r deltafig`). Here, the carbon isotope discrimination associated with g~m~ accounted for the majority of $\Delta$~o~ (69.7&pm;0.4 %) and varied little across measurement temperatures, leaf types, or warming treatments.

@@

## Variation in CO~2~ drawdown
More open stomata in shade leaves, possible as a result of a non-limiting supply of leaf K, lead to a significant increase in C~i~ compared to sun leaves (Figure `r ciccfig`a). As increases in A in shade leaves were associated with higher rates of g~s~ the C~i~ was similar than under low light but still greater than sun leaves. As the drop in CO~2~ for intercellular spaces to the cholorplast, C~i~-C~c~, measures the coordination between g~m~ and A [@caemmerer2014temperature] an increase in this drawdown was detected in shade leaves at high light (Figure `r ciccfig`b,c).  This was a direct result of the increase in g~m~ and its subsequent direct effect on A, when compared to both sun and shade leaves that are acclimated to their local light environment. CO~2~ drawdown from C~a~ to C~i~ and C~i~ to C~c~ were both relatively stable across the range of temperatures measured and gas exchange campaigns (Figure `r get2fig`c and `r get1fig`c, respectively).

@@

\
Photosynthesis in trees is also limited by the ability to supply water to the upper canopy. Ultimately, the ability of a trees hydraulic architecture to supply water to foliage across increasing pathlengths affects productivity and survival [@sellin2008effects]. Using a two-leaf model Peltoniemi et al. [-@peltoniemi2012co] theorizes that optimal N distribution will be proportional to light distribution only if leaf K is also optimally distributed. In this study, variation in leaf N distribution and A rates were not associated with subsequent changes in leaf K between sun and shade leaves. As both sun and shade leaves were acclimated to their local light environment this sub-optimal distribution of water supply has the potential to diminish the C return for the whole canopy.

\
Unexpected higher rates of g~s~ led to decreased instantaneous WUE in shade leaves throughout the experiment. Additionally, consistently higher leaf $\delta$^13^C in shade leaves suggests that this pattern was constant across leaf lifespans. From a canopy perspective this pattern in WUE initially appears to be detrimental to C gain as A in sun leaves was characterized by low rates of g~s~ and low C~i~.  As whole canopy C gain integrates the efficiency of all leaves, this begs the question of why shade leaves maintained a lower WUE while the photosynthetic efficiency of sun leaves appears at least partially constrained by water supply.

@@

\
It is possible that sustaining higher g~s~, at the cost of larger transpiration losses, is a strategy to increase the utilization efficiency of sunflecks. Rapid increases in leaf temperature with sunflecks has the potential to inhibit C gain [@leakey2003high]. It has been suggested that transpiration-induced cooling is effective in avoiding sunfleck induced heat damage, however leaves need to keep stomata open in the shade for this to occur [@schymanski2013stomatal]. Here, shade leaves were able to increase A to rates equivalent or even outperforming sun leaves under identical light intensity. Whether more open stomata is a strategy to reduce stomatal response time or to decrease thermal load on shade leaves in trees is still unknown. Unfortunately, empirical studies that assess photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@leakey2002relative] and understory plants [@chazdon1991importance; @brantley2009contribution]. Thus, our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy (see De Pury & Farquhar, [-@de1997simple]).

@@

\
Furthermore, our findings have important implications for process based models that predict canopy C gain from rates of leaf photosynthesis. Although the distribution of resources, such as N and water, define leaf photosynthetic capacity it is the physiological behaviour of individual leaves which actually determine C gain.

The observed physiological behaviour of leaves in the shade, at both low and high irradiance, suggests that current theories of leaf optimal behaviour are incomplete. Furthermore, g~m~ must now be considered as a dynamic process in tree growth models that cannot be simply parameterized.  Additional empirical data across multiple tree species are needed to determine both the mechanisms and the capacity of g~m~ to rapidly increase CO~2~ drawdown. To improve our ability to predict whole canopy C gain future research should prioritize the incorporation of both sun and shade leaf physiology, which may be optimized differently.

These results should not be entirely unexpected as shade leaves experience interchanging periods of sun and shade, despite lower mean PPFD [@pearcy1990sunflecks].

@@

**Table 1**. *Eucalyptus tereticornis* leaf traits and the distribution of resources between full sun and shade leaves under ambient and elevated temperature treatments. Leaf mass per area, leaf N~a~, $\delta$^13^C, $\Psi$~pd~, $\Psi$~l~ and K values represent treatment mean (&pm; 1 standard error) across measurement campaigns (n=6). Values of Vc~max~ and J~max~ are treatment mean(standard error) from AC~i~ curves measured in each chamber at saturating light. Different letters represent significant differences between treatments. The P value represents the overall difference between leaf types and warming treatments.

@@

\
**Figure 4**. The response of leaf photosynthesis rates to stomatal conductance (a) and mesophyll conductance (b) for sun leaves and shade leaves at both low and high light with elevated and ambient temperature treatments.  The response of shade leaf physiology to high light was recorded once both photosynthesis and stomatal conductance were stable in the gas exchange leaf cuvette (ca. 25min). Lines represent either smoothed regressions from a generalized additive model fit (a) or linear model fits (b). Grey areas are approximately 95% confidence intervals from the mean.

@@

\
**Figure 6**. (a) Response of instantaneous water use efficiency (ITE) to leaf vapor pressure deficit (VPD) for sun leaves and shade leaves at both low and high light with elevated and ambient temperature treatments. (b) The relationship between leaf $\delta$^13^C and leaf N~a~ for sun leaves and shade leaves at low light. For (a) VPD is the leaf to air pressure difference inside the gas exchange cuvette and lines represent predictions from the optimal ITE model with a g1 value for each leaf type and treatment. For (b) the dashed line represents the significant linear model fit for all leaves with a marginal and conditional r^2^ of 0.41 and 0.45, respectively.

@@

\
**Figure S1**. Photosynthetic CO~2~ response (AC~c~) curves for sun and shade leaves at elevated and ambient temperature treatments. C~c~ values were predicted with mesophyll conductance, thus curves represent chloroplastic photosynthetic parameters at 25°C and at saturating light (1800 $\mu$mols m^-1^ s^-1^).

\
**Figure S2**. Response of photosynthesis (a), mesophyll conductance (b) and C~i~-C~c~ to leaf temperature for sun leaves and shade leaves at low and high light.  Shaded symbols represents each monthly measurement campaign.

\
**Figure S3**. Response of vapor pressure deficit (a), stomatal conductance (b) and C~a~-C~i~ to leaf temperature for sun leaves and shade leaves at low and high light.  Shaded symbols represents each monthly measurement campaign.

@@

```{r Agmgs, fig.width=6, fig.height=8, echo=FALSE,warning=FALSE,message=FALSE,dev="png", dpi=400, dev.args=list(type="cairo")}
source("master_scripts/Agmgs.R")
```

@@

@@

# Supporting Information {.unnumbered}

@@

**Figure S1**.

```{r gasex_temp1, fig.width=6, fig.height=10, echo=FALSE,warning=FALSE,dev="png", dpi=400, dev.args=list(type="cairo")}
source("master_scripts/gasex_temp.R")
```

**Figure S2**.

```{r gasex_temp2, fig.width=6, fig.height=10, echo=FALSE,warning=FALSE,dev="png", dpi=400, dev.args=list(type="cairo")}
source("master_scripts/gasex_temp2.R")
```
